---
title: "Homework 1 - readable and efficient R code"
author: "Qinghua Li (730305170)"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1 - "function-alize" this code

Read over the code below and perform the following:

* Wrap it into a function `foobar0` which has arguments `x` and `z`
  and which returns the vector `x` at the end of the following code.
* Rewrite this into a function `foobar` which is easier to read, by
  reducing repetitive code. E.g. `foobar` might call a function to
  check the input, and another function to perform the three lines of
  computation. 
* Check that the two versions produce the same output using the
  function `all.equal`.

```{r}
foobar0 <- function(x, z){
  if (sum(x >= .001) < 1) {
    stop("step 1 requires 1 observation(s) with value >= .001")
  }
  
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- sin(r) + .01
  
  if (sum(x >= .002) < 2) {
    stop("step 2 requires 2 observation(s) with value >= .002")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 2 * sin(r) + .02
  
  if (sum(x >= .003) < 3) {
    stop("step 3 requires 3 observation(s) with value >= .003")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 3 * sin(r) + .03
  
  if (sum(x >= .004) < 4) {
    stop("step 4 requires 4 observation(s) with value >= .004")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 4 * sin(r) + .04
  return(x)
}
```


```{r}
foobar <- function(x, z) {
  check_input <- function(x, threshold, step) {
    if (sum(x >= threshold) < step) {
      stop(paste0("step ", step, " requires ", step, 
                  " observation(s) with value >= ", threshold))
    }
  }

  compute_step <- function(x, z, multiplier, offset) {
    fit <- lm(x ~ z)
    r <- fit$residuals
    x <- multiplier * sin(r) + offset
    return(x)
  }

  check_input(x, .001, 1)
  x <- compute_step(x, z, 1, .01)

  check_input(x, .002, 2)
  x <- compute_step(x, z, 2, .02)

  check_input(x, .003, 3)
  x <- compute_step(x, z, 3, .03)

  check_input(x, .004, 4)
  x <- compute_step(x, z, 4, .04)

  return(x)
}
```

```{r}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
all.equal(foobar(x, z), foobar0(x, z))
```


# Question 2 - vectorize this code and benchmark

* Take the following function `f0` and rewrite it as a function `f`,
  which is faster and easier to read, by removing the loop of `i` from
  1 to `m`.
* Benchmark `f` and `f0` using `microbenchmark`. How much faster is `f`?

```{r}
n <- 30
p <- 50
p2 <- 25
m <- 1000
set.seed(1)
x <- matrix(rnorm(n*p),nrow=n,ncol=p)
b <- matrix(rnorm(m*p),nrow=m,ncol=p)
a <- matrix(rnorm(m*p2),nrow=m,ncol=p2)
f0 <- function(x,b,a) {
  out <- numeric(0)
  for (i in seq_len(m)) {
    bb <- b[i,]
    aa <- a[i,]
    out <- c(out, sum(x %*% bb) + sum(aa))
  }
  out
}
```

```{r, warning=FALSE}
f1 <- function(x, b, a){
  colSums(x %*% t(b) + rowSums(a))
}


microbenchmark::microbenchmark(
  f0 = f0(x, b, a),
  f1 = f1(x, b, a),
  times = 100
)
```


# Question 3 - build a faster t-test

* Rewrite the following function `getT0` which computes `m`
  two-sample t-tests (equal variance) between two groups as a function
  `getT`, which is faster by using vectorized operations over the `m`
  sets of observations. (There are functions in R packages, such as
  `genefilter::rowttests` which will quickly perform this operation,
  but I want you to write your own function using simple R functions
  like `rowSums`, etc.)
* Benchmark `getT` and `getT0`. How much faster is `getT`?

```{r}
m <- 400
n <- 50
little.n <- n/2
set.seed(1)
x <- matrix(rnorm(m*n), nrow=m,ncol=n)
f <- gl(2,little.n)
getT0 <- function(x, f) {
  ts <- sapply(seq_len(m), function(i) t.test(x[i,] ~ f, var.equal=TRUE)$statistic)
  unname(ts)
}
```


```{r}
getT <- function(x, f){
  group1 <- rowMeans(x[, f == levels(f)[1]])
  group2 <- rowMeans(x[, f == levels(f)[2]])
  n1 <- sum(f == levels(f)[1])
  n2 <- sum(f == levels(f)[2])
  s1 <- rowSums((x[, f == levels(f)[1]] - group1)^2) / (n1 - 1)
  s2 <- rowSums((x[, f == levels(f)[2]] - group2)^2) / (n2 - 1)
  pooled_se <- sqrt((s1 / n1) + (s2 / n2))
  (group1 - group2) / pooled_se
}
```


```{r}
microbenchmark::microbenchmark(
  getT0 = getT0(x, f),
  getT = getT(x, f),
  times = 100
)
```



