---
title: "HW 8 - MCMC"
author: "Qinghua Li"
date: "3/31/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importance and Rejection Sampling

Consider finding $\sigma^2 = E[X^2]$ when $X$ has the density that is proportional to $q(x) =\exp({−|x|^3/3})$.  Write the form of the integral pertaining to this expectation below, assuming the normalizing constant for $q(x)$ is $c$.

Solution:


## Find the estimate for $\sigma^2$ using importance sampling.  

Write the expression for the integral using an importance sampling scheme, specifying your choice for the importance sampling density $g$ and reexpress the integral in terms of $X^2$, $g(x)$ and $q(x)$.  Then, write the form of the approximation of this integrand based on the importance samples drawn from $g$. Then, carry out this approximation of the expectation in the code below. 

$$
\begin{align}
\sigma^2 &= E(X^2) \\
         &= \int_{-\infty}^{+\infty} cx^2 q(x)dx \\
         &= \int_{-\infty}^{+\infty} cx^2 exp(-|x|^3/3)dx \\
         &= \int_{-\infty}^{+\infty} cx^2 \frac {exp(-|x|^3/3)}{\frac{1}{\sqrt{2\pi}}exp{(-x^2/2)}} \frac{1}{\sqrt{2\pi}}exp{(-x^2/2)}dx \\
         &\approx \frac {\sum_{i=1}^{M} \frac {x_{i}^2exp(-|x_{i}|^3/3)}{\frac{1}{\sqrt{2\pi}}exp(-x_{i}^2/2)}} {\sum_{i=1}^{M} \frac {exp(-|x_{i}|^3/3)}{\frac{1}{\sqrt{2\pi}}exp(-x_{i}^2/2)}}
\end{align}
$$


```{r}
## Solution: write any helper functions here (q(x), etc)
q <- function(x){
  exp(-abs(x)^3/3)
}

## End Solution

# set M
M = 10^5

## Solution: place primary code evaluating the expectation here with importance sampling
x <- rnorm(M)
intergral <- sum(x^2*q(x)/dnorm(x))/sum(q(x)/dnorm(x))
intergral
```


## Repeat the procedure using rejection sampling. 

Similar to before, write out the form of the integral and then its approximation, this time based on sampling from the target density $q(x)$.  Then, carry out this approximation of the expectation in the code below. 

Hint:  We just need to show that $e$ exceeds the target density everywhere, and that the distribution $g$ that we draw samples from only differs from $e$ by a constant $\alpha$.  


```{r}
# set M
M = 10^5

## Solution: place primary code evaluating the expectation here with rejection sampling
maxit = M * 10^4
i = index = 1
x = rep(NA, M)
alpha = q(1)/exp(-1/2)

while (i < maxit & index <= M) {
    # sample from the normal distribution
    y = rnorm(1)
    # sample from the uniform distribution
    u = runif(1)
    i = i + 1
    if (u <= q(y) / (alpha * exp(-y^2/2))) {
      x[index] = y
      index = index + 1
    }
}
int_rej = mean(x^2)
int_rej
## End Solution
```